### 2017-02-08 2017mergingmerges/section/01-Introduction.tex
Edits to S2

In this implementation the @group@ combinator filters out consecutive duplicates, while @merge@ combines two sorted arrays such that the output remains sorted. There are a few interesting things about this example. 
Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ reads values from @arr1@ or @arr2@ depends on the values in the arrays themselves. If all the values in @arr1@ are smaller than the values in @arr2@, then @merge@ will pull values from @arr1@ before pulling from @arr2@, and vice versa.
Secondly, although @arr1@ occurs twice in the program, at runtime we only want to read through the elements of this array once. To achieve this the compiled program must coordinate between the two uses of @arr1@, so when the implementation of @uniques@ requires a new value the implementation of @union@ is also ready to receive a new value, and vice versa.
Although the program is expressed in terms of bulk \emph{arrays}, at runtime we want it to execute using some version of \emph{streaming data flow}, in order to save memory bandwidth, avoid repeated computation, and eliminate the need for intermediate buffers such as @merged@.

@@

In our new system we view the program as a concurrent process network, where each combinator is a separate process, and the streams are communications channels between the processes. Each combinator is expressed as a restricted, sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing to an output channel. The fusion transform then takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.

@@

\ben{Add a few more sentences on related work. Explain how this work extends the old flow fusion paper. It is not short-cut fusion like Oleg's recent work. We are not in the same space as Fortran style array fusion transformations like polyhedral}
\amos{I'm not convinced anybody reading this cares about Fortran or polyhedral}

>>>>

In this implementation the @group@ combinator filters out consecutive duplicates, while @merge@ combines two sorted arrays so that the output remains sorted. There are a few interesting things about this example. 
Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ reads values from @arr1@ or @arr2@ depends on the values in the arrays themselves. If all the values in @arr1@ are smaller than the values in @arr2@, then @merge@ will pull values from @arr1@ before pulling from @arr2@, and vice versa.
Secondly, although @arr1@ occurs twice in the program, at runtime we only want to read through the elements of this array once. To achieve this the compiled program must coordinate between the two uses of @arr1@, so when the implementation of @uniques@ requires a new value the implementation of @union@ is also ready to receive a new value, and vice versa.
Although the program is expressed in terms of bulk \emph{arrays}, at runtime we want it to execute using some version of \emph{streaming data flow}, in order to save memory bandwidth, avoid repeated computation, and eliminate the need for intermediate buffers such as @merged@.

@@

In our new system we view the program as a concurrent process network, where each combinator is a separate process, and the streams are communication channels between the processes. Each combinator is expressed as a restricted, sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing to an output channel. The fusion transform then takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.

@@

\ben{Add a few more sentences on related work. Explain how this work extends the old flow fusion paper. It is not short-cut fusion like Oleg's recent work. We are not in the same space as Fortran style array fusion transformations like polyhedral}
\amos{I'm not convinced anybody reading this cares about Fortran or polyhedral}
\ben{I agree. I meant that we should cite related work to say that's also called ``fusion'', but in a completely different domain.}

### 2017-02-08 2017mergingmerges/section/XX-Processes.tex
Edits to S2

Each combinator is defined as a stream process and expressed in terms of a simple imperative program with a local heap. The process pulls data from an aribrary number if input streams and pushes data to at least one output stream. \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.

\subsection{Grouping}
The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Figure~\ref{fig:Process:Group}

\begin{figure}
\begin{alltt}
 group 
   = \(\lambda\) (s1: Stream a) (s2: Stream a). 
     \(\nu\) (first: Bool)  (last: a) (value: a) (L0..L3: Label).
\end{alltt}
\begin{code}
     process
     { ins:    { s1 }
     , outs:   { s2 }
     , heap:   { first = True, last = 0, value = 0 }
     , label:  L0
     , instrs: { L0 = pull s1 value                    L1 {}
               , L1 = case (first || (last /= value))  L2 {}  L3 {}
               , L2 = push s2 value                    L3 { last = value, first = False }
               , L3 = drop s1                          L0 {} } }
\end{code}

\medskip

\begin{alltt}
 merge
   = \(\lambda\) (s1: Stream a) (s2: Stream a) (s3: Stream a). 
     \(\nu\) (a: a)  (b: a) (value: a) (L0..L3: Label) (FIXME).
\end{alltt}
\begin{code}
     process
     { ins:    { s1, s2 }
     , outs:   { s3 }
     , heap:   { a = 0, b = 0 }
     , label:  L0
     , instrs: { L0 = pull s1 a  L1 {}
               , L1 = pull s2 b  C0 {}
               , C0 = case (a < b)  A0 {}  B0 {}
               , A0 = push s3 a A1 {}
               , A1 = drop s1    A2 {}
               , A2 = pull s1  a C0 {}
               , B0 = push merged b B1 {}
               , B1 = drop s2       B2 {}
               , B2 = pull s2  b    C0 {} } }
\end{code}
\caption{Grouping and Merging}
\label{fig:Process:Group}
\end{figure}

The @group@ combinator has two parameters, @s1@ and @s2@ which are the bind the input and output streams respectively. The \emph{nu-binders} \mbox{$\nu$ @(first: Bool)@...} indicate that each time we instantiate the @group@ combinator we should create fresh names for @first@, @last@ and so on that do not conflict with other instantiations.

@@

\subsection{Merging}

Finally, we will look at the process for @merge@.
Merge takes two input streams, @file1@ and @file2@, and produces the output @merged@.
The two heap variables @a@ and @b@ are for storing the currently read values from both inputs.
Again, they are initialised to $0$, but the initial values will not be used.
Merge requires more complicated control flow than before.
Instructions @L0@ and @L1@ initialise the process by reading the first values from each stream, then move to @C0@.
Instruction @C0@ checks which value is smaller: if the value @a@ read from @file1@ is smaller, it moves to @A0@; otherwise it moves to @B0@.
Instructions @A0@, @A1@ and @A2@ push @file1@'s value to the output, drop it, and pull another one before moving back to @C0@ to compare again.
Instructions @B0@, @B1@ and @B2@ are equivalent, except performing on @file2@ instead.

As this process merges infinite streams, evaluating with finite streams will block waiting for more input once the end of either stream is reached.
This is not an issue for infinite streams, and we use examples involving finite streams are much easier to describe.
We address finite streams in~\S\ref{s:Finite}.
Executing this process with the inputs $[1, 4]$ and $[2, 3, 100]$ will produce the first four elements $[1, 2, 3, 4]$ as expected, but note that the $100$ is missing because the process is blocked waiting for the next value from @file1@.
\amos{I don't think this is necessary}

\subsection{Fusion}
We are now ready to take our @group@ and @merge@ processes and fuse them together. 
We create a new process with inputs @file1@ and @file2@ and outputs @uniques@ and @merged@.
In order to make the fused process easier to understand, we have performed some minor optimisations on the output of the fusion algorithm described in~\S\ref{s:Fusion}, such as removing @jump@s and merging the two variables @value@ and @a@ into a single variable @value_a@.
Rather than explaining the process in detail, we simply note that the fused process is an interleaved scheduling of the two, starting with executing @group@, then executing @merge@, and going back to @group@ whenever pulling from @file1@ again.
The labels @L0@-@L2@ and @A1@-@A4@ are much the same as those in @group@, while @C0@, @A0@-@A2@ and @B0@-@B2@ are similar to those in @merge@.

@@

\subsection{Process definitions}

>>>>

Each combinator defines a stream process which is expressed in terms of a simple imperative program with a local heap. The process pulls data from an aribrary number if input streams and pushes data to at least one output stream. \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.

% -----------------------------------------------------------------------------
\subsection{Grouping}
\begin{figure}
\begin{alltt}
 group 
   = \(\lambda\) (s1: Stream a) (s2: Stream a). 
     \(\nu\) (first: Bool)  (last: a) (value: a) (L0..L3: Label).
\end{alltt}
\begin{code}
     process
     { ins:    { s1 }
     , outs:   { s2 }
     , heap:   { first = True, last = 0, value = 0 }
     , label:  L0
     , instrs: { L0 = pull s1 value                    L1 {}
               , L1 = case (first || (last /= value))  L2 {}  L3 {}
               , L2 = push s2 value                    L3 { last = value, first = False }
               , L3 = drop s1                          L0 {} } }
\end{code}
\caption{The group combinator}
\label{fig:Process:Group}
\end{figure}

The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Figure~\ref{fig:Process:Group}. The @group@ combinator has two parameters, @s1@ and @s2@ which bind the input and output streams respectively. The \emph{nu-binders} like \mbox{$\nu$ @(first: Bool)@...} indicate that each time we instantiate the @group@ combinator we should create fresh names for @first@, @last@ and so on that do not conflict with other instantiations.

@@

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure}
\begin{alltt}
 merge
   = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat) (s3: Stream Nat). 
     \(\nu\) (x1: Nat)  (x2: Nat) (L0..L3: Label) (FIXME).
\end{alltt}
\begin{code}
     process
     { ins:    { s1, s2 }
     , outs:   { s3 }
     , heap:   { x1 = 0, x2 = 0 }
     , label:  L0
     , instrs: { L0 = pull s1 x1     L1 {}
               , L1 = pull s2 x2     C0 {}
               , C0 = case (x1 < x2) A0 {}  B0 {}
               , A0 = push s3 x1     A1 {}
               , A1 = drop s1        A2 {}
               , A2 = pull s1 x1     C0 {}
               , B0 = push s3 x2     B1 {}
               , B1 = drop s2        B2 {}
               , B2 = pull s2  x2    C0 {} } }
\end{code}
\caption{The merge combinator}
\label{fig:Process:Merge}
\end{figure}

The definition of the @merge@ combinator which merges two input streams is given in Figure~\ref{fig:Process:Merge}. The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@. The two heap variables @x1@ and @x2@ are used to store the currently read values from each input. The forms of the instructions used to define this combinator are the same as the @group@ combinator in the previous section. We only need the four basic @pull@, @case@, @push@ and @drop@ instructions.

As this process merges infinite streams, if we execute it using a finite prefix then the final state will be an intermediate one that may not yet have pushed all available output. For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output. The machine will arrive at instruction @L?@ which blocks waiting for the next value to be pulled from @arr1@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}

% The instructions at @L0@ and @L1@ initialise the process by reading the first values from each stream, then move to @C0@.
% Instruction @C0@ checks which value is smaller: if the value @a@ read from @file1@ is smaller, it moves to @A0@; otherwise it moves to @B0@.
% Instructions @A0@, @A1@ and @A2@ push @file1@'s value to the output, drop it, and pull another one before moving back to @C0@ to compare again.
% Instructions @B0@, @B1@ and @B2@ are equivalent, except performing on @file2@ instead.

% -----------------------------------------------------------------------------
\subsection{Fusion}
We are now ready to take our @group@ and @merge@ processes and fuse them together. 
We create a new process with inputs @file1@ and @file2@ and outputs @uniques@ and @merged@.
In order to make the fused process easier to understand, we have performed some minor optimisations on the output of the fusion algorithm described in~\S\ref{s:Fusion}, such as removing @jump@s and merging the two variables @value@ and @a@ into a single variable @value_a@.
Rather than explaining the process in detail, we simply note that the fused process is an interleaved scheduling of the two, starting with executing @group@, then executing @merge@, and going back to @group@ whenever pulling from @file1@ again.
The labels @L0@-@L2@ and @A1@-@A4@ are much the same as those in @group@, while @C0@, @A0@-@A2@ and @B0@-@B2@ are similar to those in @merge@.

@@

\clearpage{}
\subsection{Process definitions}
